Modelling multi-scale, state-switching functional data with hidden Markov models: Supplement B (simulation study)

### 2021-12-07 Report/Sections/appendix.tex
Update Report

We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$,

@@

\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
\]
%
\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
\]
and
\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
\]
Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a gamma distribution gives the desired values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.

@@

\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.\]
%
where

\begin{align*}
P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
\end{align*}
%
and $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$. The likelihood corresponding to the fine-scale model is

>>>>

We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse DFT of $\hat{Y}^*_{t,\tilde t^*}$,

@@

\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
\]
%
\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right),
\]
and
\[
    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
\]
Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a gamma distribution gives the desired values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.

@@

\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N\]
%
where

\begin{align*}
P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big]
\end{align*}
%
and $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$. The likelihood corresponding to the fine-scale model is

### 2021-12-07 Report/Sections/results.tex
Update Report

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., DeRuiter et al., 2017; Isojunno et al.,2017)\nocite{DeRuiter:2017,Isojunno:2017}. For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook salmon \citep{Williams:2009,Noren:2011,Wright:2017}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter and shallower dives that likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer and deeper and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth greater than 30 $m$, while \citet{Wright:2017}, in a study of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 $m$. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
Dive type 2 could also be associated with behaviours such as socializing, which can take place several meters below the surface \citep{Tennessen:2019b}.

@@

\[\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.745 & 0.253 & 0.002 \\
0.080 & 0.869 & 0.052 \\
0.000 & 0.229 & 0.771
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.886 & 0.114 & 0.000 \\
0.150 & 0.815 & 0.036 \\
0.000 & 0.225 & 0.775
\end{pmatrix},\]

$\hat \delta^{*(1)} = \left(0.202, 0.649, 0.149\right)$, and $\hat \delta^{*(2)} = \left( 0.531, 0.405, 0.064 \right)$ for dive types 1 and 2.
In summary, about 86\% of dives are short dives of type 1. The whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of which describe common bouts of short resting dives before a longer, more-energy-intensive deep dive.
Further, this killer whale is in the less-active subdive state 1 only 20\% of the time during a dive of type 1, compared to 53\% of the time during a dive of type 2. Less-active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to greater depths and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 4 of Supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.

>>>>

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., DeRuiter et al., 2017; Isojunno et al., 2017)\nocite{DeRuiter:2017,Isojunno:2017}. For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook salmon \citep{Williams:2009,Noren:2011,Wright:2017}. 
Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter and shallower dives that likely reflect resting, travelling, and, to a lesser extent, searching for prey.
Dive type 2 is longer and deeper and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth greater than 30 $m$, while \citet{Wright:2017}, in a study of killer whales in Johnstone Strait, found that most prey captures occur at depths deeper than 100 $m$. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
Dive type 2 could also be associated with behaviours such as socializing, which can take place several meters below the surface \citep{Tennessen:2019b}.

@@

\[\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.745 & 0.253 & 0.002 \\
0.080 & 0.869 & 0.052 \\
0.000 & 0.229 & 0.771
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.886 & 0.114 & 0.000 \\
0.150 & 0.815 & 0.036 \\
0.000 & 0.225 & 0.775
\end{pmatrix},\]
%
$\hat \delta^{*(1)} = \left(0.202, 0.649, 0.149\right)$, and $\hat \delta^{*(2)} = \left( 0.531, 0.405, 0.064 \right)$ for dive types 1 and 2.
In summary, about 86\% of dives are short dives of type 1. The whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of which describe common bouts of short resting dives before a longer, more-energy-intensive deep dive.
Further, this killer whale is in the less-active subdive state 1 only 20\% of the time during a dive of type 1, compared to 53\% of the time during a dive of type 2. Less-active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to greater depths and holding their breath for long periods \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 4 of Supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.

### 2021-12-07 Report/Sections/simulation.tex
Update Report

\[\Gamma = \begin{pmatrix} 0.847 & 0.153 \\ 0.914 & 0.086 \end{pmatrix}.\]
%
Dive duration is gamma distributed and the coarse-scale emission parameters are $\mu^{(1)} = 27.34$ $s$, $\sigma^{(1)} = 10.96$ $s$, $\mu^{(2)} = 127.55$ $s$, and $\sigma^{(2)} = 63.89$ $s$.
%
After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with transition probability matrices

@@

We calculate the maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using an optimization procedure similar to that in the case study.

For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows. Each test data set consists of a sequence of 100 dives and is created from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$. To assess the coarse-scale hidden state prediction, we estimate $p_t(i \mid y,\z) = \Pr(X_t=i \mid Y=y,\Z=\z)$ for $i=1,2$ and $t=1,\ldots,100$ using the test set observations $(y,\z)$ and the training set maximum likelihood estimates. These estimates are found using the forward--backward algorithm \citep{Zucchini:2016}. We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the average dive decoding accuracy for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t \mid y,\z)/100$. We then report the average of this measure over the 500 training/test data set pairs. 
Analogously, to assess fine-scale state prediction, we estimate $p^*_{t,\tilde t^*}(i^* \mid y,\z) = \Pr(\tilde X^*_{t,\tilde t^*}=i \mid Y=y,\Z=\z)$ for $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, and $t=1,\ldots,100$ using the test set observations, the training set maximum likelihood estimates, and the forward--backward algorithm. Denoting the true fine-scale state realizations from the test data set by $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall average subdive decoding accuracy as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*} \mid y,\z)$ across all simulated test data sets, dives, and windows. The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward--backward algorithm.

>>>>

\[\Gamma = \begin{pmatrix} 0.847 & 0.153 \\ 0.914 & 0.086 \end{pmatrix}.\]
%
Dive duration is gamma distributed and the coarse-scale emission parameters are $\mu^{(1)} = 27.34$ $s$, $\sigma^{(1)} = 10.96$ $s$, $\mu^{(2)} = 127.55$ $s$, and $\sigma^{(2)} = 63.89$ $s$.
%
After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each 2 $s$ segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with the transition probability matrices

@@

We calculate the maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using an optimization procedure similar to that in the case study.
%
For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows. Each test data set consists of a sequence of 100 dives and is created from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$. To assess the coarse-scale hidden state prediction, we estimate $p_t(i \mid y,\z) = \Pr(X_t=i \mid Y=y,\Z=\z)$ for $i=1,2$ and $t=1,\ldots,100$ using the test set observations $(y,\z)$ and the training set maximum likelihood estimates. These estimates are found using the forward--backward algorithm \citep{Zucchini:2016}. We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the average dive decoding accuracy for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t \mid y,\z)/100$. We then report the average of this measure over the 500 training/test data set pairs. 
Analogously, to assess fine-scale state prediction, we estimate $p^*_{t,\tilde t^*}(i^* \mid y,\z) = \Pr(\tilde X^*_{t,\tilde t^*}=i \mid Y=y,\Z=\z)$ for $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, and $t=1,\ldots,100$ using the test set observations, the training set maximum likelihood estimates, and the forward--backward algorithm. Denoting the true fine-scale state realizations from the test data set by $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall average subdive decoding accuracy as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*} \mid y,\z)$ across all simulated test data sets, dives, and windows. The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward--backward algorithm.
